---
title: "Hospital Readmission Prediction - Diabetes Patients"
author: "Noelia Caba Martin"
date: "2026-03-19"
output: html_document
---

## Introduction
**The goal of this project is to identify patients at high risk of 
30-day readmission in order to support early intervention strategies, 
reduce healthcare costs, and improve patient outcomes.**
Hospital readmission — when a patient is discharged and returns within 
30 days — is a major challenge in healthcare. It indicates inadequate 
recovery, increases healthcare costs, and is associated with worse 
patient outcomes.

In the United States, the Centers for Medicare & Medicaid Services (CMS) 
penalizes hospitals with high readmission rates, making prediction of 
at-risk patients a high-priority clinical and business problem.

**Why diabetes patients?**
Diabetic patients have particularly high readmission rates due to:
- Complexity of disease management
- Multiple comorbidities
- Medication adjustments during hospitalization

**Research question:** Can we predict 30-day hospital readmission in 
diabetic patients from clinical and administrative data?

The main objectives are:

- Perform comprehensive data cleaning of real hospital records
- Engineer clinically meaningful features
- Build a predictive model for 30-day readmission
- Evaluate model performance with appropriate metrics
- Identify key clinical predictors of readmission

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(tidyverse)
library(caret)
library(randomForest)
library(pROC)
library(MLmetrics)
library(ggplot2)
library(corrplot)
```

## Data Loading and Initial Exploration
```{r load-data}
# Load dataset
hospital <- read.csv("diabetic_data.csv", 
                     stringsAsFactors = FALSE,
                     na.strings = c("?", "Unknown", ""))

# Initial exploration
dim(hospital)
str(hospital)
head(hospital, 3)
```

The dataset contains **101,766 hospital encounters** and **50 variables** 
covering patient demographics, clinical measurements, diagnoses and 
medications.

Key observations:

- **`readmitted`** has three levels: `NO`, `<30` (readmitted within 
30 days) and `>30` (readmitted after 30 days). For this analysis, 
we define the outcome as **readmission within 30 days** (`<30` = Yes, 
all others = No)
- **`weight`** has extensive missing data and will be removed
- **`age`** is coded as ranges and requires recoding
- **23 medication variables** are coded as No/Steady/Up/Down

## Data Cleaning

This dataset requires extensive cleaning due to its real-world clinical 
origin. Key steps include:

1. **Remove irrelevant variables** — IDs, variables with excessive missingness
2. **Create binary outcome** — readmission within 30 days
3. **Recode age** — convert ranges to numeric midpoints
4. **Handle medications** — simplify to binary (changed vs not changed)
5. **Remove duplicate patients** — keep only first encounter per patient
6. **Filter invalid records** — remove deceased/hospice patients

```{r data-cleaning}
hospital_clean <- hospital %>%
  
  # Remove variables with excessive missingness or irrelevant
  select(-weight, -payer_code, -encounter_id) %>%
  
  # Keep only first encounter per patient
  distinct(patient_nbr, .keep_all = TRUE) %>%
  select(-patient_nbr) %>%
  
  # Remove patients discharged to hospice or who died
  filter(!discharge_disposition_id %in% c(11, 13, 14, 19, 20, 21)) %>%
  
  # Create binary outcome: readmitted within 30 days
  mutate(readmitted_30 = ifelse(readmitted == "<30", "Yes", "No")) %>%
  mutate(readmitted_30 = as.factor(readmitted_30)) %>%
  select(-readmitted) %>%
  
  # Recode age to numeric midpoints
  mutate(age = case_when(
    age == "[0-10)"  ~ 5,
    age == "[10-20)" ~ 15,
    age == "[20-30)" ~ 25,
    age == "[30-40)" ~ 35,
    age == "[40-50)" ~ 45,
    age == "[50-60)" ~ 55,
    age == "[60-70)" ~ 65,
    age == "[70-80)" ~ 75,
    age == "[80-90)" ~ 85,
    age == "[90-100)" ~ 95
  )) %>%
  
  # Simplify medication variables to binary
  mutate(across(c(metformin, insulin, glipizide, glyburide,
                  pioglitazone, rosiglitazone),
                ~ifelse(. == "No", 0, 1))) %>%
  
  # Convert gender to binary
  mutate(gender = ifelse(gender == "Male", 1, 0)) %>%
  
  # Remove remaining character variables except race
  select(where(is.numeric), race, medical_specialty,
         change, diabetesMed, A1Cresult, readmitted_30) %>%
  
  mutate(medical_specialty = as.factor(medical_specialty),
       race = as.factor(race),
       change = as.factor(change),
       diabetesMed = as.factor(diabetesMed),
       A1Cresult = as.factor(A1Cresult)) %>%
  
  # Remove rows with missing values
  drop_na()

# Check results
dim(hospital_clean)
table(hospital_clean$readmitted_30)
prop.table(table(hospital_clean$readmitted_30))
```

After cleaning, the dataset contains **35,394 unique patient encounters** 
and **25 variables**. Key cleaning steps applied:

- Removed variables with excessive missingness (weight, payer_code)
- Kept only the **first encounter per patient** to avoid data leakage
- Removed patients discharged to hospice or who died (not at risk 
of readmission)
- Recoded age ranges to numeric midpoints
- Simplified medication variables to binary
- Created binary outcome: **readmission within 30 days**

**Class distribution:**
- Not readmitted within 30 days: 32,301 (91.3%)
- Readmitted within 30 days: 3,093 (8.7%)

The significant class imbalance reflects real clinical data — early 
readmission is relatively rare but clinically critical.

## Exploratory Data Analysis

```{r eda}
# 1. Readmission by age
hospital_clean %>%
  group_by(age, readmitted_30) %>%
  summarise(n = n()) %>%
  mutate(pct = n / sum(n)) %>%
  filter(readmitted_30 == "Yes") %>%
  ggplot(aes(x = age, y = pct)) +
  geom_col(fill = "#E64B35", alpha = 0.8) +
  geom_smooth(se = FALSE, color = "#2E9FDF") +
  theme_minimal() +
  labs(title = "30-Day Readmission Rate by Age",
       x = "Age", y = "Readmission Rate")

# 2. Readmission by time in hospital
hospital_clean %>%
  group_by(time_in_hospital, readmitted_30) %>%
  summarise(n = n()) %>%
  mutate(pct = n / sum(n)) %>%
  filter(readmitted_30 == "Yes") %>%
  ggplot(aes(x = time_in_hospital, y = pct)) +
  geom_col(fill = "#2E9FDF", alpha = 0.8) +
  theme_minimal() +
  labs(title = "30-Day Readmission Rate by Time in Hospital",
       x = "Days in Hospital", y = "Readmission Rate")

# 3. Readmission by number of diagnoses
hospital_clean %>%
  group_by(number_diagnoses, readmitted_30) %>%
  summarise(n = n()) %>%
  mutate(pct = n / sum(n)) %>%
  filter(readmitted_30 == "Yes") %>%
  ggplot(aes(x = number_diagnoses, y = pct)) +
  geom_col(fill = "#00A087", alpha = 0.8) +
  theme_minimal() +
  labs(title = "30-Day Readmission Rate by Number of Diagnoses",
       x = "Number of Diagnoses", y = "Readmission Rate")
```

The exploratory analysis reveals several clinically meaningful patterns:

**Age:** Readmission rates increase with age, peaking in older patients 
(70-90 years). This is consistent with the higher disease burden and 
complexity of care in elderly diabetic patients.

**Time in hospital:** Longer hospitalizations are associated with higher 
readmission rates, likely reflecting more severe disease episodes that 
require continued management after discharge.

**Number of diagnoses:** Patients with more diagnoses (higher comorbidity 
burden) show substantially higher readmission rates, confirming that 
clinical complexity is a key risk factor.

## Model Training

**Why Random Forest?**

Random Forest was selected due to its:
- Robustness to noisy and incomplete clinical data
- Ability to capture non-linear relationships between variables
- Built-in variable importance for clinical interpretability
- Strong performance in published readmission prediction studies

**Why 5-fold CV with SMOTE?**

Given the extreme class imbalance (8.7% positive cases), SMOTE was 
applied within each fold to generate synthetic minority samples without 
data leakage. 5-fold CV provides a robust, unbiased estimate of 
real-world performance.
We train a Random Forest model with cross-validation and SMOTE to 
handle the class imbalance (91.3% No vs 8.7% Yes), following the 
best practices established in the previous projects.

```{r model-training}
# Split into training (80%) and test (20%)
set.seed(123)
train_index <- createDataPartition(hospital_clean$readmitted_30,
                                   p = 0.8, list = FALSE)
train_data <- hospital_clean[train_index, ]
test_data <- hospital_clean[-train_index, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")

# Define cross-validation with SMOTE
cv_control <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  sampling = "smote"
)

# Train Random Forest
set.seed(123)
rf_model <- train(
  readmitted_30 ~ .,
  data = train_data,
  method = "rf",
  metric = "ROC",
  trControl = cv_control,
  tuneGrid = expand.grid(mtry = c(3, 5, 7)),
  ntree = 300
)

print(rf_model)
```

The cross-validated model achieved **ROC-AUC = 0.625** with mtry = 5 
as the optimal parameter.

The results reveal the fundamental challenge of this dataset: with only 
8.7% positive cases, the model struggles to balance sensitivity and 
specificity. High sensitivity (99.9%) comes at the cost of near-zero 
specificity, reflecting the extreme class imbalance.

This is a common finding in real-world hospital readmission prediction 
studies — ROC-AUC values between 0.60-0.70 are typical in the literature 
for this type of data.

## Model Evaluation

We evaluate the model on the held-out test set using multiple metrics, 
paying special attention to PR-AUC and sensitivity given the clinical 
context — missing a high-risk patient has serious consequences.

```{r model-evaluation}
# Predictions on test set
rf_pred <- predict(rf_model, test_data)
rf_prob <- predict(rf_model, test_data, type = "prob")[, "Yes"]

# Confusion matrix
conf_matrix <- confusionMatrix(rf_pred, test_data$readmitted_30,
                                positive = "Yes")
print(conf_matrix)

# ROC-AUC
roc_obj <- pROC::roc(test_data$readmitted_30, rf_prob)
cat("ROC-AUC:", round(pROC::auc(roc_obj), 3), "\n")

# PR-AUC
pr_obj <- precrec::evalmod(
  scores = rf_prob,
  labels = ifelse(test_data$readmitted_30 == "Yes", 1, 0)
)
pr_aucs <- precrec::auc(pr_obj)
cat("PR-AUC:", round(pr_aucs[pr_aucs$curvetypes == "PRC", "aucs"], 3), "\n")

# F1 Score
f1 <- MLmetrics::F1_Score(
  y_pred = rf_pred,
  y_true = test_data$readmitted_30,
  positive = "Yes"
)
cat("F1 Score:", round(f1, 3), "\n")

# ROC curve
plot(roc_obj,
     main = paste("ROC Curve - AUC:", round(pROC::auc(roc_obj), 3)),
     col = "#E64B35", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey")
```

The model evaluation reveals the genuine difficulty of predicting 
rare clinical events:

- **ROC-AUC = 0.627:** the model shows meaningful discriminative 
ability above random chance (0.5), consistent with published 
literature on readmission prediction (typical range: 0.60-0.70)
- **PR-AUC = 0.140 vs baseline 0.087:** the model identifies 
high-risk patients at a rate 60% better than random selection
- **Sensitivity at threshold 0.10 = ~75%:** the model successfully 
identifies approximately 3 out of 4 high-risk patients, 
significantly reducing the likelihood of missed readmissions

These results should not be interpreted as model failure — they 
reflect the **inherent complexity of clinical prediction** where 
many readmissions are driven by social and environmental factors 
not captured in administrative data.

## Threshold Tuning and Variable Importance

Given the extreme class imbalance, we tune the classification threshold 
to improve sensitivity at the cost of specificity. In a clinical screening 
context, missing a high-risk patient is more costly than a false alarm.

We also examine which variables the model considers most important for 
predicting readmission.

```{r threshold-vip}
library(vip)

# Threshold tuning
thresholds <- seq(0.05, 0.5, by = 0.05)

threshold_results <- data.frame(
  threshold = thresholds,
  sensitivity = NA,
  specificity = NA,
  f1 = NA
)

for(i in seq_along(thresholds)) {
  pred_thresh <- factor(
    ifelse(rf_prob > thresholds[i], "Yes", "No"),
    levels = c("No", "Yes")
  )
  cm <- confusionMatrix(pred_thresh, test_data$readmitted_30,
                        positive = "Yes")
  threshold_results$sensitivity[i] <- cm$byClass["Sensitivity"]
  threshold_results$specificity[i] <- cm$byClass["Specificity"]
  threshold_results$f1[i] <- MLmetrics::F1_Score(
    y_pred = pred_thresh,
    y_true = test_data$readmitted_30,
    positive = "Yes"
  )
}

# Plot threshold tuning
threshold_results %>%
  pivot_longer(-threshold, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("#E64B35", "#2E9FDF", "#00A087")) +
  theme_minimal() +
  labs(title = "Threshold Tuning",
       x = "Threshold", y = "Value", color = "Metric")

# Variable importance
vip(rf_model,
    num_features = 15,
    aesthetics = list(fill = "#E64B35", alpha = 0.8)) +
  theme_minimal() +
  labs(title = "Variable Importance - Readmission Prediction")
```

**Feature importance analysis — critical for clinical interpretability:**

The three most important predictors are:

- **Gender:** male and female diabetic patients show fundamentally 
different readmission patterns, likely reflecting differences in 
healthcare-seeking behavior, social support and disease management
- **Insulin:** changes in insulin dosage during hospitalization 
directly reflect glycemic instability — the primary driver of 
diabetic complications and readmission
- **Medication change:** patients whose medications were not adjusted 
during hospitalization may represent undertreated cases, leaving 
them at higher risk after discharge

👉 These findings suggest that **targeted medication review at 
discharge**, particularly for insulin-dependent patients, could 
reduce readmission rates.

**Threshold Tuning:**

At the default threshold (0.5), the model detects no readmission cases. 
Lowering the threshold to 0.05-0.10 achieves sensitivity of ~75%, 
though at the cost of lower specificity (~30%).

In a clinical screening context, a threshold of **0.10** would be 
recommended — identifying approximately 75% of high-risk patients 
while maintaining acceptable false positive rates.

## Conclusions

This analysis demonstrates the challenges and opportunities of predicting 
30-day hospital readmission in diabetic patients using real-world 
clinical data.

**Key findings:**

1. After cleaning 101,766 hospital records, **35,394 unique patient 
encounters** were retained for analysis. The cleaning process addressed 
duplicate patients, deceased/hospice cases, and missing values — 
reflecting the complexity of real hospital data.

2. The dataset shows extreme class imbalance (**8.7% readmission rate**), 
consistent with real clinical data. This makes prediction genuinely 
difficult and highlights the importance of appropriate evaluation metrics.

3. The Random Forest model achieved **ROC-AUC = 0.627**, consistent 
with published literature on hospital readmission prediction (typical 
range: 0.60-0.70). This moderate performance reflects the inherent 
difficulty of predicting complex clinical outcomes.

4. **Gender, insulin management and medication changes** were identified 
as the strongest predictors of readmission, consistent with clinical 
knowledge about diabetes management.

5. **Threshold tuning** revealed that a threshold of 0.10 achieves 
~75% sensitivity — clinically preferable for screening despite lower 
specificity.

**Business and clinical impact:**

In the United States, hospitals with high readmission rates face 
financial penalties from CMS. With approximately **35,000 diabetic 
patients** in this dataset:

- The model identifies ~75% of high-risk patients at a threshold of 0.10
- This represents approximately **2,300 patients** who could receive 
targeted interventions
- Even a **10% reduction in readmissions** would translate into 
significant cost savings and improved patient outcomes

This demonstrates how predictive models can directly support 
**clinical decision-making and hospital resource allocation**.

**Clinical implications:**

This model could serve as a **preliminary screening tool** to flag 
high-risk patients before discharge, enabling targeted interventions 
such as:
- Enhanced discharge planning
- Earlier follow-up appointments
- Medication reconciliation
- Patient education programs

**Limitations:**

- Moderate predictive performance (ROC-AUC = 0.627) limits clinical utility
- Important predictors like HbA1c trends and social determinants 
of health were not available
- Model requires external validation before clinical deployment
- Class imbalance remains a major challenge despite SMOTE

**Future directions:**

- Incorporate additional clinical variables (HbA1c trajectory, 
social support, medication adherence)
- Explore gradient boosting methods (XGBoost, LightGBM)
- Apply cost-sensitive learning with explicit readmission costs
- Validate on an independent hospital dataset
- Develop a Shiny dashboard for clinical use